---
title: "Covid-open-data"
author: "David Henderson"
date: "06/06/2020"
output: 
  html_document:
    theme: journal
    highlight: haddock
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.width = 9, fig.height = 6)
```

# Introduction

## Packages

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
#Get most up-to-date version of opendatascot
#remotes::install_github("datasciencescotland/opendatascot", force = TRUE)
library(janitor)
library(lubridate)
library(opendatascot) 
library(here)
library(curl)
#library(phsmethods)   #remotes::install.github("Health-SocialCare-Scotland/phsmethods")

`%nin%` <- negate(`%in%`)

theme_set(theme_minimal(base_family = "Roboto", base_size = 20) +
            theme(panel.grid.minor = element_blank(),
                  axis.title.y = element_text(margin = margin(0, 20, 0, 0)),
                  axis.title.x = element_text(margin = margin(20, 0, 0, 0)),
                  plot.caption = element_text(colour = "#AAAAAA")))

colour_palette <- c("#011f4b", "#03396c", "#005b96", "#6497b1", "#b3cde0")

options(scipen = 10)
```

## Data

### Main data

Scrape deaths dataset - note I am selecting Scotland only here - it is possible to split by LA or Health Board...

(*Note* issue with API where dates are not being coerced correctly and information is lost. Therefore downloading from the csv link for now)

```{r, warning=FALSE, message=FALSE}

# need a lookup file for date to week number. 

#cv_data_main <- ods_dataset("deaths-involving-coronavirus-covid-19",
#                            #return only national data
#                           geography = "sc") %>% 
#  select(-refArea, -measureType)



cv_data_main <- read_csv("https://statistics.gov.scot/downloads/cube-table?uri=http%3A%2F%2Fstatistics.gov.scot%2Fdata%2Fdeaths-involving-coronavirus-covid-19") %>% 
  clean_names() %>% 
  filter(feature_code == "S92000003") %>% 
  mutate(date_code = str_replace(date_code, "w/c ", ""),
         date_code = ymd(date_code)) %>% 
  filter(!is.na(date_code)) %>% 
  select(date_code, sex, age, cause_of_death, location_of_death, measurement, value, 
         -feature_code, -units) %>% 
  arrange(date_code, sex, age, cause_of_death, location_of_death, measurement)

cv_data_main 
```

```{r, eval=FALSE}
#Save data for report
feather::write_feather(cv_data_main, "derived_data/cv_data_main.feather")
```



### Management info

There is a lot of information in this extract. I am filtering to just include delayed discharges and care home information for Scotland as a whole. You can see all the available info with the `ods_structure("coronavirus-covid-19-management-information")` command. 


```{r}
cv_data_manage <- ods_dataset("coronavirus-covid-19-management-information",
                              variable = 
                                c("delayed-discharges",
      "adult-care-homes-cumulative-number-that-have-reported-a-suspected-covid-19-case",
      "adult-care-homes-proportion-that-have-reported-a-suspected-covid-19-case",
      "adult-care-homes-cumulative-number-that-have-reported-more-than-one-suspected-covid-19-case",
      "adult-care-homes-number-with-current-suspected-covid-19-cases",
      "adult-care-homes-proportion-with-current-suspected-covid-19-cases",
      "adult-care-homes-cumulative-number-of-suspected-covid-19-cases",
      "adult-care-homes-daily-number-of-new-suspected-covid-19-cases",
      "adult-care-homes-number-of-staff-reported-as-absent",
      "adult-care-homes-adult-care-homes-which-submitted-a-return",
      "adult-care-homes-response-rate",
      "adult-care-homes-total-number-of-staff-in-adult-care-homes-which-submitted-a-return",
      "adult-care-homes-staff-absence-rate"),
      geography = "sc") %>% 
  select(-refArea)
cv_data_manage
```

```{r, eval=FALSE}
#Save data for report
feather::write_feather(cv_data_manage, "derived_data/cv_data_manage.feather")
```


### Population estimates

API is not good for population data - it keeps on timing out on me plus most recent figures are for 2014. 
The NHS Scotland open data file is also very large so will read in with `fread()` directly from url.

```{r}
#pop_est_raw <- data.table::fread("https://www.opendata.nhs.scot/dataset/7f010430-6ce1-4813-b25c-f7f335bdc4dc/resource/c505f490-c201-44bd-abd1-1bd7a64285ee/download/dz2011-pop-est_30082019.csv") %>% 
#  clean_names %>% 
#  filter(year == "2018") %>% 
#  select(-year)

#feather::write_feather(pop_est_raw, "derived_data/pop_est_raw.feather")
pop_est_raw <- feather::read_feather("derived_data/pop_est_raw.feather")


pop_est_raw %>% 
  mutate(`0 years` = rowSums(pop_est_raw[,c(5)]),
         `1-14 years` = rowSums(pop_est_raw[,c(15:19)]),
         `15-44 years` = rowSums(pop_est_raw[,c(20:49)]),
         `45-64 years` = rowSums(pop_est_raw[,c(50:69)]),
         `65-74 years` = rowSums(pop_est_raw[,c(70:79)]),
         `75-84 years` = rowSums(pop_est_raw[,c(80:89)]),
         `85 years and over` = rowSums(pop_est_raw[,c(90:95)])) %>% 
  select(`0 years`:`85 years and over`) %>%
  slice(1:2) %>% 
  summarise(`0 years` = sum(`0 years`),
         `1-14 years` = sum(`1-14 years`),
         `15-44 years` = sum(`15-44 years`),
         `45-64 years` = sum(`45-64 years`),
         `65-74 years` = sum(`65-74 years`),
         `75-84 years` = sum(`75-84 years`),
         `85 years and over` = sum(`85 years and over`)) %>% 
  pivot_longer(cols = `0 years`:`85 years and over`, 
               names_to = "age", 
               values_to = "pop_est_2018") -> pop_est_1
```


```{r, eval=FALSE}
#Save data for report
feather::write_feather(pop_est_1, "derived_data/pop_est_1.feather")
```


```{r}
pop_est_raw %>% 
  mutate(`18-64-years` = rowSums(pop_est_raw[,c(23:69)]),
         `65-74-years` = rowSums(pop_est_raw[,c(70:79)]),
         `75-84-years` = rowSums(pop_est_raw[,c(80:89)]),
         `85-years-and-over` = rowSums(pop_est_raw[,c(90:95)])) %>% 
  select(`18-64-years`:`85-years-and-over`) %>%
  slice(1:2) %>% 
  summarise(`18-64-years` = sum(`18-64-years`),
         `65-74-years` = sum(`65-74-years`),
         `75-84-years` = sum(`75-84-years`),
         `85-years-and-over` = sum(`85-years-and-over`)) %>% 
  pivot_longer(cols = `18-64-years`:`85-years-and-over`, 
               names_to = "age", 
               values_to = "pop_est_2018") -> pop_est_2
feather::write_feather(pop_est_2, "derived_data/pop_est_2.feather")
```



### Care home characteristics


```{r}
ods_dataset("care-homes-demography",
            refPeriod = "2017",
            measureType = "count",
            sex = "all",
            mainClientGroupInCareHome = "all-adults",
            geography = "sc") %>% 
  select(age, value) %>% 
  mutate(value = as.numeric(value)) -> df

df %>% 
  bind_rows(
    df %>% 
      filter(age %in% c("85-94-years", "95-years-and-over")) %>% 
      summarise_if(is.double, sum) %>% 
      mutate(age = "85-years-and-over")
  ) %>% 
  filter(age %nin% c("85-94-years", "95-years-and-over")) -> ch_age_groups

feather::write_feather(ch_age_groups, "derived_data/ch_age_groups.feather")
```


# Plots

## Figure 1

Simple count and % of deaths.

```{r, fig.width=12, fig.height=9}
cv_data_main %>% 
  filter(sex == "All" & location_of_death == "All" & cause_of_death == "COVID-19 related") %>% 
  filter(age != "All") %>% 
  group_by(age) %>% 
  summarise(n = sum(value)) %>%
  mutate(pct = round(n/sum(n), 3)) %>% 
  ggplot(aes(age, n)) +
  geom_col(fill = colour_palette[3]) +
  geom_text(aes(label = scales::percent(pct, accuracy = 0.1), y = n + 80),
            size = 6) +
  scale_y_continuous(limits = c(0, 2000)) +
  labs(x = "",
       y = "", 
       title = "Deaths Associated with COVID-19 by Age Group",
       subtitle = "16th March 2020 to 25th May 2020",       
       caption = "Source: National Records of Scotland\nDeaths where COVID-19 mentioned on the death certificate") -> fig_1

fig_1
```

```{r, eval=FALSE}
ggsave("plots/20200609_fig_1.png", fig_1, width = 12, height =9, dpi = 300)
```


## Figure 2


```{r, fig.width=12, fig.height=9}
cv_data_main %>% 
  filter(sex == "All" & location_of_death == "All" & cause_of_death == "COVID-19 related") %>% 
  filter(age != "All") %>% 
  group_by(age) %>% 
  summarise(n = sum(value)) %>% 
  left_join(., pop_est_1) %>% 
  mutate(risk = n/pop_est_2018,
         av_risk = sum(n)/sum(pop_est_2018),
         relative_risk = risk/av_risk) %>% 
  ggplot(aes(age, relative_risk)) +
  geom_col(fill = colour_palette[3]) +
  geom_text(aes(label = round(relative_risk,1), 
                y = relative_risk + 1), size = 6) +
  labs(title = "Risk of dying from COVID-19 relative to Scottish average",
       subtitle = "16th March 2020 to 25th May 2020",
       x = "",
       y = "Relative Risk",
       caption = "Source:National Records of Scotland\nand opendata.nhs.scot (2018 mid-year population estimates)")
```

```{r, eval=FALSE}
ggsave("plots/20200609_fig_2.png", fig_2, width = 12, height =9, dpi = 300)
```


## Figure 3

```{r, fig.width=12, fig.height=9}
ch_age_groups %>%
  left_join(., pop_est_2) %>% 
  mutate(pct = value/pop_est_2018) %>% 
  ggplot(aes(age, pct)) +
  geom_col(fill = colour_palette[3]) +
  geom_text(aes(label = scales::percent(pct, accuracy = 0.1), y = pct + 0.01),
            size = 6) +
  scale_y_continuous(limits = c(0, 0.2),
                     labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Percentage of Specified Age groups Resident in Care Homes in Scotland",
       subtitle = "2017",
       x = "",
       y = "",
       caption = "Source: Scottish Care Home Census via statistics.gov.scot\nand opendata.nhs.scot (2018 mid-year population estimates)") -> fig_3
fig_3
```


```{r, eval=FALSE}
ggsave("plots/20200609_fig_3.png", fig_3, width = 12, height =9, dpi = 300)
```

## Figure 4



# Session Info

```{r}
devtools::session_info()
```


