---
title: "Covid-open-data"
author: "David Henderson"
date: "06/06/2020"
output: 
  html_document:
    theme: journal
    highlight: haddock
    df_print: paged
    code_folding: hide
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.width = 12, fig.height = 9)
```

# Introduction

## Packages

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
#Get most up-to-date version of opendatascot
#remotes::install_github("datasciencescotland/opendatascot", force = TRUE)
library(janitor)
library(lubridate)
library(opendatascot) 
library(here)
library(curl)
library(ggthemes)
library(readxl)


`%nin%` <- negate(`%in%`)

theme_set(theme_minimal(base_family = "Roboto", base_size = 20) +
            theme(panel.grid.minor = element_blank(),
                  axis.title.y = element_text(margin = margin(0, 20, 0, 0)),
                  axis.title.x = element_text(margin = margin(20, 0, 0, 0)),
                  plot.caption = element_text(colour = "#AAAAAA"),
                  plot.margin = margin(3,15,3,3,"mm")))
          
options(scipen = 10)
```

## Data

### Main data

Scrape deaths dataset - note I am selecting Scotland only here - it is possible to split by LA or Health Board...


```{r, warning=FALSE, message=FALSE}
#Download data via API
cv_data_main <- 
ods_dataset("deaths-involving-coronavirus-covid-19",
                            #return only national data
                           geography = "sc") %>% 
  select(-refArea, -measureType) %>% 
  clean_names() %>% 
  mutate(causeofdeath = str_replace(
           causeofdeath,
           "all-causes-average-of-corresponding-week-over-previous-5-years",
           "all-cause_5yr_average"),
         causeofdeath = factor(causeofdeath),  
         age = factor(age, levels = c("all", "0-years", "1-14-years", "15-44-years",
                                      "45-64-years", "65-74-years", "75-84-years",
                                      "85-years-and-over")),
         value = as.integer(value)) 
cv_data_main
```

```{r, eval=FALSE}
#Save data for report
feather::write_feather(cv_data_main, "derived_data/cv_data_main.feather")
```

### Weekly

```{r}
# Create a look-up table for week numbers to dates
week_lookup <- tibble(
  week_number = 1:52,
  week_date = seq(ymd(20191230), ymd(20201221), by = "1 week"))

cv_data_main %>% 
  filter(ref_period != "2020") %>% 
  rename(week_number = ref_period) %>% 
  mutate(week_number = str_replace(week_number, "2020-", ""),
         week_number = as.integer(week_number)) %>% 
  left_join(., week_lookup) %>% 
  select(week_date, week_number, everything()) -> cv_data_weekly
cv_data_weekly
```

```{r, eval=FALSE}
#Save data for report
feather::write_feather(cv_data_weekly, "derived_data/cv_data_weekly.feather")
```


### Management info

There is a lot of information in this extract. I am filtering to just include delayed discharges and care home information for Scotland as a whole. You can see all the available info with the `ods_structure("coronavirus-covid-19-management-information")` command. 


```{r}
cv_data_manage <- 
  ods_dataset("coronavirus-covid-19-management-information",
                              variable = 
                                c("delayed-discharges",
      "adult-care-homes-cumulative-number-that-have-reported-a-suspected-covid-19-case",
      "adult-care-homes-proportion-that-have-reported-a-suspected-covid-19-case",
      "adult-care-homes-cumulative-number-that-have-reported-more-than-one-suspected-covid-19-case",
      "adult-care-homes-number-with-current-suspected-covid-19-cases",
      "adult-care-homes-proportion-with-current-suspected-covid-19-cases",
      "adult-care-homes-cumulative-number-of-suspected-covid-19-cases",
      "adult-care-homes-daily-number-of-new-suspected-covid-19-cases",
      "adult-care-homes-number-of-staff-reported-as-absent",
      "adult-care-homes-adult-care-homes-which-submitted-a-return",
      "adult-care-homes-response-rate",
      "adult-care-homes-total-number-of-staff-in-adult-care-homes-which-submitted-a-return",
      "adult-care-homes-staff-absence-rate"),
      geography = "sc") %>% 
  select(-refArea) %>% 
  clean_names() %>% 
  rename(date = ref_period) %>% 
  mutate(date = ymd(date))
cv_data_manage
```

```{r, eval=FALSE}
#Save data for report
feather::write_feather(cv_data_manage, "derived_data/cv_data_manage.feather")
```


### Population estimates

API is not good for population data - it keeps on timing out on me plus most recent figures are for 2014. 
The NHS Scotland open data file is also very large so will read in with `fread()` directly from url.

Because this is a big file I am filtering it to 2018 and then saving as a feather file. This means it is much quicker to re-run the rmarkdown frequently. Commenting out the original code but can be rerun at any time.

```{r}
#pop_est_raw <- data.table::fread("https://www.opendata.nhs.scot/dataset/7f010430-6ce1-4813-b25c-f7f335bdc4dc/resource/c505f490-c201-44bd-abd1-1bd7a64285ee/download/dz2011-pop-est_30082019.csv") %>% 
#  clean_names %>% 
#  filter(year == "2018") %>% 
#  select(-year)

#feather::write_feather(pop_est_raw, "derived_data/pop_est_raw.feather")
pop_est_raw <- feather::read_feather("derived_data/pop_est_raw.feather")


pop_est_raw %>% 
  mutate(`0-years` = rowSums(pop_est_raw[,c(5)]),
         `1-14-years` = rowSums(pop_est_raw[,c(15:19)]),
         `15-44-years` = rowSums(pop_est_raw[,c(20:49)]),
         `45-64-years` = rowSums(pop_est_raw[,c(50:69)]),
         `65-74-years` = rowSums(pop_est_raw[,c(70:79)]),
         `75-84-years` = rowSums(pop_est_raw[,c(80:89)]),
         `85-years-and-over` = rowSums(pop_est_raw[,c(90:95)])) %>% 
  select(`0-years`:`85-years-and-over`) %>%
  slice(1:2) %>% 
  summarise(`0-years` = sum(`0-years`),
         `1-14-years` = sum(`1-14-years`),
         `15-44-years` = sum(`15-44-years`),
         `45-64-years` = sum(`45-64-years`),
         `65-74-years` = sum(`65-74-years`),
         `75-84-years` = sum(`75-84-years`),
         `85-years-and-over` = sum(`85-years-and-over`)) %>% 
  pivot_longer(cols = `0-years`:`85-years-and-over`, 
               names_to = "age", 
               values_to = "pop_est_2018") -> pop_est_1
```


```{r, eval=FALSE}
#Save data for report
feather::write_feather(pop_est_1, "derived_data/pop_est_1.feather")
```


```{r}
pop_est_raw %>% 
  mutate(`18-64-years` = rowSums(pop_est_raw[,c(23:69)]),
         `65-74-years` = rowSums(pop_est_raw[,c(70:79)]),
         `75-84-years` = rowSums(pop_est_raw[,c(80:89)]),
         `85-years-and-over` = rowSums(pop_est_raw[,c(90:95)])) %>% 
  select(`18-64-years`:`85-years-and-over`) %>%
  slice(1:2) %>% 
  summarise(`18-64-years` = sum(`18-64-years`),
         `65-74-years` = sum(`65-74-years`),
         `75-84-years` = sum(`75-84-years`),
         `85-years-and-over` = sum(`85-years-and-over`)) %>% 
  pivot_longer(cols = `18-64-years`:`85-years-and-over`, 
               names_to = "age", 
               values_to = "pop_est_2018") -> pop_est_2
feather::write_feather(pop_est_2, "derived_data/pop_est_2.feather")
```



### Care home characteristics


```{r}
ods_dataset("care-homes-demography",
            refPeriod = "2017",
            measureType = "count",
            sex = "all",
            mainClientGroupInCareHome = "all-adults",
            geography = "sc") %>% 
  select(age, value) %>% 
  mutate(value = as.numeric(value)) -> df

df %>% 
  bind_rows(
    df %>% 
      filter(age %in% c("85-94-years", "95-years-and-over")) %>% 
      summarise_if(is.double, sum) %>% 
      mutate(age = "85-years-and-over")
  ) %>% 
  filter(age %nin% c("85-94-years", "95-years-and-over")) -> ch_age_groups

feather::write_feather(ch_age_groups, "derived_data/ch_age_groups.feather")
```


### Delayed Discharges

```{r}
dd <-
  ods_dataset("delayed-discharges-bed-days", 
              geography = "sc") %>% 
  clean_names() %>% 
  select(-ref_area, -measure_type) %>% 
  rename(date = ref_period) %>% 
  mutate(date = ymd(date, truncated = 2),
         value = as.double(value))

dd
```

### Excess deaths

Unfortunately the related stats published by NRS aren't on the SG API, so will still need to download and wrangle these files directly...sigh

```{r, warning=FALSE, message=FALSE}
temp_1 <- tempfile()
temp_2 <- tempfile()
source <- "https://www.nrscotland.gov.uk/files//statistics/covid19/covid-deaths-data-week-27.zip"

temp_1 <- curl_download(url = source, destfile = temp_1, quiet = FALSE)
unzip(temp_1, exdir = temp_2)

ch_death_covid <- read_csv(file.path(temp_2,"covid-deaths-data-week-27_Table 1 - COVID deaths.csv"), skip = 3)
ch_death_all <- read_csv(file.path(temp_2, "covid-deaths-data-week-27_Table 2 - All deaths.csv"), skip = 3)
ch_death_average <- read_csv("https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-2015-2019.csv", skip = 2)
```


#### By location


```{r,fig.width=9, fig.height=6}
loc_names <- c("Care Home", "Home and other non institution", "Hospital",
          "Other institution")

#Wrangle Average deaths
ch_death_average %>% 
  slice(2:6, 9:13, 16:20, 23:27) %>% 
  mutate(location = rep(loc_names, each = 5, times = 1)) %>% 
  select(year = `Week number2`, location, everything(), -`53`) %>% 
  pivot_longer(cols = `1`:`52`, names_to = "week_number",
               values_to = "n_deaths") %>% 
  group_by(location, week_number) %>%
  mutate(min_deaths = min(n_deaths),
         max_deaths = max(n_deaths),
         mean_deaths = mean(n_deaths)) %>% 
  distinct(location, week_number, .keep_all = TRUE) %>% 
  select(-year, -n_deaths) %>% 
  ungroup %>% 
  mutate(week_number = as.integer(week_number)) -> main_tab

#Wrangle all deaths
ch_death_all %>% 
  select(location = X2, everything(), -`Week beginning`, -X30:-X66) %>% 
  slice(85:88) %>%  
  mutate_at(vars(`30-Dec-19`:`29-Jun-20`), as.numeric) %>% 
  pivot_longer(cols = `30-Dec-19`:`29-Jun-20`, names_to = "date",
               values_to = "deaths_all_2020") %>% 
  mutate(date = dmy(date),
         week_number = rep(1:27, each = 1, times = 4),
         location = rep(loc_names, each = 27, times = 1)) %>% 
  select(-date) %>% 
  left_join(main_tab, .) -> main_tab


#Wrangle Covid deaths
ch_death_covid %>% 
  select(location = X2, everything(), -`Week beginning`,
         -`Year to Date`:-X45) %>% 
  slice(83:86) %>%  
  pivot_longer(cols = `30-Dec-19`:`29-Jun-20`, names_to = "date",
               values_to = "deaths_covid_2020") %>% 
  mutate(date = dmy(date),
         week_number = rep(1:27, each = 1, times = 4),
         location = rep(loc_names, each = 27, times = 1)) %>% 
  select(-date) %>% 
  left_join(main_tab, .) %>% 
  mutate(deaths_nonCovid_2020 = deaths_all_2020 - deaths_covid_2020,
         location = fct_relevel(location, "Hospital")) %>% 
  left_join(., week_lookup) -> main_tab
```

```{r, echo=FALSE, eval=FALSE}
feather::write_feather(main_tab, "derived_data/main_tab.feather")
```


#### By age

```{r}
#Wrangle Average deaths
ch_death_average %>% 
  slice(2:6, 9:13, 16:20, 23:27) %>% 
  mutate(location = rep(loc_names, each = 5, times = 1)) %>% 
  select(year = `Week number2`, location, everything(), -`53`) %>% 
  pivot_longer(cols = `1`:`52`, names_to = "week_number",
               values_to = "n_deaths") %>% 
  group_by(location, week_number) %>%
  mutate(min_deaths = min(n_deaths),
         max_deaths = max(n_deaths),
         mean_deaths = mean(n_deaths)) %>% 
  distinct(location, week_number, .keep_all = TRUE) %>% 
  select(-year, -n_deaths) %>% 
  ungroup %>% 
  mutate(week_number = as.integer(week_number)) -> main_tab

#Wrangle all deaths
ch_death_all %>% 
  select(location = X2, everything(), -`Week beginning`, -X27:-X63) %>% 
  slice(85:88) %>%  
  mutate_at(vars(`30-Dec-19`:`8-Jun-20`), as.numeric) %>% 
  pivot_longer(cols = `30-Dec-19`:`8-Jun-20`, names_to = "date",
               values_to = "deaths_all_2020") %>% 
  mutate(date = dmy(date),
         week_number = rep(1:24, each = 1, times = 4),
         location = rep(loc_names, each = 24, times = 1)) %>% 
  select(-date) %>% 
  left_join(main_tab, .) -> main_tab


#Wrangle Covid deaths
ch_death_covid %>% 
  select(location = X2, everything(), -`Week beginning`,
         -`Year to Date`:-X42) %>% 
  slice(83:86) %>%  
  pivot_longer(cols = `30-Dec-19`:`8-Jun-20`, names_to = "date",
               values_to = "deaths_covid_2020") %>% 
  mutate(date = dmy(date),
         week_number = rep(1:24, each = 1, times = 4),
         location = rep(loc_names, each = 24, times = 1)) %>% 
  select(-date) %>% 
  left_join(main_tab, .) %>% 
  mutate(deaths_nonCovid_2020 = deaths_all_2020 - deaths_covid_2020,
         location = fct_relevel(location, "Hospital")) %>% 
  left_join(., week_lookup) -> main_tab
rm(list = c("ch_death_average", "ch_death_covid", "ch_death_all"))
```





# Plots

## Figure 1

Simple count and % of deaths.

```{r, fig.width=12, fig.height=9}
max_date <- max(cv_data_weekly$week_date) %>% format("%d-%b-%Y")

cv_data_main %>% 
  filter(sex == "all" & locationofdeath == "all" & ref_period == "2020",
           causeofdeath == "covid-19-related" & age != "all") %>% 
  group_by(age) %>% 
  summarise(n = sum(value)) %>%
  mutate(pct = round(n/sum(n), 3)) %>% 
  ggplot(aes(age, n)) +
  geom_col(fill = economist_pal()(1)) +
  geom_text(aes(label = scales::percent(pct, accuracy = 0.1), y = n + 80),
            size = 6) +
  scale_y_continuous(limits = c(0, 2000)) +
  theme(axis.text.x = element_text(size = 14)) +
  labs(x = "",
       y = "", 
       title = "Deaths Associated with COVID-19 by Age Group",
       subtitle = glue::glue("16-March-2020 to {max_date}"),       
       caption = "Source: National Records of Scotland\nDeaths where COVID-19 mentioned on the death certificate") -> fig_1

fig_1
```

```{r, eval=FALSE}
ggsave("plots/20200710_fig_1.png", fig_1, width = 12, height =9, dpi = 300)
```


## Figure 2


```{r, fig.width=12, fig.height=9}


cv_data_main %>% 
  filter(sex == "all" & locationofdeath == "all" & 
           causeofdeath == "covid-19-related" &
           ref_period == "2020" & age != "all") %>% 
  group_by(age) %>% 
  summarise(n = sum(value)) %>% 
  left_join(., pop_est_1) %>% 
  mutate(risk = n/pop_est_2018,
         av_risk = sum(n)/sum(pop_est_2018),
         relative_risk = risk/av_risk) %>% 
  ggplot(aes(age, relative_risk)) +
  geom_col(fill = economist_pal()(1)) +
  geom_text(aes(label = round(relative_risk,1), 
                y = relative_risk + 1), size = 6) +
  scale_y_continuous(limits = c(0, 20)) +
  theme(axis.text.x = element_text(size = 14)) +
  labs(title = "Risk of dying from COVID-19 relative to Scottish average",
       subtitle = glue::glue("16-March-2020 to {max_date}"),
       x = "",
       y = "Relative Risk",
       caption = "Source:National Records of Scotland\nand opendata.nhs.scot (2018 mid-year population estimates)")
```

```{r, eval=FALSE}
ggsave("plots/20200609_fig_2.png", fig_2, width = 12, height =9, dpi = 300)
```


## Figure 3

```{r, fig.width=12, fig.height=9}
ch_age_groups %>%
  left_join(., pop_est_2) %>% 
  mutate(pct = value/pop_est_2018) %>% 
  ggplot(aes(age, pct)) +
  geom_col(fill = economist_pal()(1)) +
  geom_text(aes(label = scales::percent(pct, accuracy = 0.1), y = pct + 0.01),
            size = 6) +
  scale_y_continuous(limits = c(0, 0.2),
                     labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Percentage of Specified Age groups Resident in Care Homes in Scotland",
       subtitle = "2017",
       x = "",
       y = "",
       caption = "Source: statistics.gov.scot- Care Homes-Demographic characteristics of residents\nand opendata.nhs.scot- 2018 mid-year population estimates") -> fig_3
fig_3
```


```{r, eval=FALSE}
ggsave("plots/20200609_fig_3.png", fig_3, width = 12, height =9, dpi = 300)
```

## Figure 4

```{r, fig.width=12, fig.height=6}
cv_data_manage %>% 
  filter(variable %in% 
           c("adult-care-homes-proportion-that-have-reported-a-suspected-covid-19-case")) %>%
  mutate(value = as.integer(value)) %>% 
  ggplot(aes(ymd(date), value)) +
  geom_line(colour = economist_pal()(1)) +
  scale_x_date(breaks = scales::pretty_breaks()) +
  scale_y_continuous(labels = scales::percent_format(scale = 1),
                     limits = c(0, 80)) +
  labs(title = "Proportion of care homes reporting suspected COVID-19 cases",
       subtitle = "2020",
       x = "",
       y = "",
       caption = "Source: statistics.gov.scot- COVID-19 Management information")
```

## Figure 5

```{r}
other <- cv_data_weekly %>% 
  filter(sex == "all" & age == "all" & causeofdeath == "covid-19-related" &
           locationofdeath == "other-institution") %>% 
  summarise(sum = sum(value)) %>% 
  pull(.)

cv_data_weekly %>% 
  filter(sex == "all" & age == "all" & causeofdeath == "covid-19-related" &
           locationofdeath %nin% c("all", "other-institution")) %>% 
  ggplot(aes(week_date, value, group = locationofdeath, 
             colour = locationofdeath)) +
  geom_point(size = 1.8) +
  geom_line(size = 1.4) +
  scale_colour_wsj(labels = c("Home/non-institution", "Care Home", "Hospital"),
                   guide = guide_legend(reverse = TRUE)) +
  scale_y_continuous(limits = c(0, 400)) +
  scale_x_date(breaks = scales::pretty_breaks(n = 12)) +
  theme(legend.position = "top",
        axis.text.x = element_text(size = 12)) +
  labs(title = "Weekly deaths involving COVID-19 in Scotland by setting",
       x = "Week beginning",
       y = "",
       colour = "",
       caption = glue::glue("Source:NRS Scotland- deaths where COVID-19 mentioned on the death certificate\nExcludes other institutional place of death n = {other}"))
```

## Figure 6


```{r}
cv_data_weekly %>% 
  filter(week_number >13 & sex == "all" & age == "all" & causeofdeath == "covid-19-related" &
           locationofdeath %nin% c("other-institution", "all")) %>% 
  group_by(week_date) %>% 
  mutate(pct = round(value/sum(value) * 100, 0)) %>% 
  arrange(week_date, locationofdeath) %>% 
  mutate(value = as.double(value),
         label_value = case_when(
           locationofdeath == "hospital" ~ value/2,
           locationofdeath == "care-home" ~ value/2 + lead(value),
           locationofdeath == "-non-institution" & value >22 ~ lead(value, 1) + 
             lead(value, 2) + value/2,
           locationofdeath == "-non-institution" & value <=22 ~ lead(value, 1) + 
             lead(value, 2) + (value +20)),
         label_colour = case_when(
           sum(value) < label_value ~ "black",
           sum(value) > label_value ~ "white")) %>% 
  ungroup -> stack_data

stack_data %>% 
  ggplot(aes(week_date, value, fill = locationofdeath)) +
  geom_col(position = "stack") +
  geom_text(aes(x = week_date, y = label_value, label = paste0(pct,"%")),
            size = 5, color = stack_data$label_colour) +
  scale_x_date(breaks = scales::pretty_breaks(n = 12)) +
  scale_y_continuous(limits = c(0, 800)) +
  scale_fill_economist(labels = c("Home/non-institution", "Care Home", "Hospital"),
                       guide = guide_legend(reverse = TRUE)) +
  theme(legend.position = "top") +
  labs(title = "Weekly Deaths from COVID-19",
       subtitle = "by setting",
       x = "Week beginning",
       y = "",
       fill = "",
       caption = glue::glue("Source:NRS Scotland- deaths where COVID-19 mentioned on the death certificate\nExcludes other institutional place of death n = {other}"))
```

## Figure 7

```{r}
cv_data_weekly %>% 
  filter(week_number >= 12 & sex == "all" & age == "all" &
           causeofdeath %in% c("covid-19-related", "all-causes") & 
           locationofdeath %nin% c("other-institution", "all")) %>%
  group_by(causeofdeath, locationofdeath) %>%
  summarise(value = sum(value)) %>% 
  pivot_wider(names_from = "causeofdeath", values_from = "value") %>% 
  mutate(pct_covid = round(`covid-19-related`/sum(`covid-19-related`)* 100,1),
         pct_all = round(`covid-19-related`/`all-causes`* 100, 1)) %>% 
  pivot_longer(c(pct_all, pct_covid), names_to = "measure", values_to = "value") %>%
  mutate(locationofdeath = factor(locationofdeath,
                                    levels = c("care-home", "hospital", "-non-institution"),
                                    labels = c("Care Home", "Hospital", "Home/non-institution")),
         measure = factor(measure, 
                          levels = c("pct_covid", "pct_all"),
                          labels = c("% of all Scottish\nCOVID-19 deaths",
                                     "% of all deaths in location\nattributable to COVID-19"))) %>% 
  ggplot(aes(locationofdeath, value, fill = measure)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = paste0(value,"%")),
            position = position_dodge(width = 1),
            vjust = -1) +
  scale_fill_economist() +
  scale_y_continuous(limits = c(0, 60),
                     labels = scales::percent_format(scale = 1)) +
  theme(legend.position = "top") +
  labs(title = "Percentage of deaths related to COVID-19",
       subtitle = glue::glue("16-March-2020 to {max_date}"),
       x = "",
       y = "",
       fill = "",
       caption = glue::glue("Source:NRS Scotland- deaths where COVID-19 mentioned on the death certificate\nExcludes other institutional place of death n = {other}"))
  
```

## Figure 9

```{r}
dd_plot <- 
  dd %>% 
  filter(reason_for_delay %nin% c("all", 
                                  "health-and-social-care-and-patient-carer-family-related-reasons") &
           age == "75-years-and-over") %>% 
  ggplot(aes(date, value, colour = reason_for_delay, group = reason_for_delay)) +
  geom_point() +
  geom_line(size = 1.2) +
  scale_colour_wsj(labels = c("Complex", "Health and Social care", "Patient/carer/family"),
                   guide = guide_legend(title.position = "top", title.hjust = 0.5)) +
  scale_x_date(breaks = scales::pretty_breaks(n = 8), date_labels = "%b-%Y") +
  scale_y_continuous(limits = c(0, 30000)) +
  theme(legend.position = "top") +
  labs(title = "Number of bed days occupied by hospital inpatients\nready for discharge in Scotland",
       subtitle = "Over 75s only",
       x = "",
       y = "Bed days",
       colour = "Delay reason",
       caption = "Source: statistics.scot.gov - Delayed discharges - monthly bed days occupied")
dd_plot
```




```{r, eval=FALSE}
ggsave("plots/20200615_dd_plot.png", dd_plot, width = 12, height =9, dpi = 300)
```


## Figure 10

```{r, fig.width=14, fig.height=10.5, warning=FALSE, message=FALSE}
ann_text_1 <- data.frame(location = 
                         factor("Hospital", 
                                levels = c("Hospital", "Care Home",
                                           "Home and other non institution",
                                           "Other")),
                       week_date = ymd(20200907), 
                       deaths = 675, 
                       label = "Shaded area shows range\nof deaths 2015-2019")

ann_text_2 <- data.frame(location = 
                         factor("Hospital", 
                                levels = c("Hospital", "Care Home",
                                           "Home and other non institution",
                                           "Other")),
                       week_date = ymd(20200907), 
                       deaths = 350, 
                       label = "text")

ann_text_3 <- data.frame(location = 
                         factor("Care Home", 
                                levels = c("Hospital", "Care Home",
                                           "Home and other non institution",
                                           "Other")),
                       week_date = ymd(20200907), 
                       deaths = 350, 
                       label = "text")

ann_text_4 <- data.frame(location = 
                         factor("Home and other non institution", 
                                levels = c("Hospital", "Care Home",
                                           "Home and other non institution",
                                           "Other")),
                       week_date = ymd(20200928), 
                       deaths = 420, 
                       label = "text",
                       linetype = "dashed")

other_inst_all_2020 <-
  main_tab %>% 
  filter(location == "Other institution") %>% 
  summarise(total_other_inst = sum(deaths_all_2020, na.rm = TRUE)) %>% 
  pull(.)


fig_10 <-
  main_tab %>% 
  filter(location %nin% "Other institution") %>% 
  pivot_longer(cols = c(deaths_all_2020, deaths_nonCovid_2020), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot() +
  geom_ribbon(aes(x = week_date,
                  ymax = max_deaths, ymin = min_deaths), fill = "#4477aa",
              alpha = 0.2) +
  geom_line(aes(x = week_date, y = deaths, colour = deaths_type)) +
  geom_line(aes(x = week_date, y = mean_deaths), linetype = "dashed",
            size = 0.2) +
  facet_wrap(~location) +
  geom_text(data = ann_text_1, 
            aes(x = week_date, y = deaths,
                label = "Shaded area shows range\nof deaths 2015-2019\ndashed line shows weekly average"),
            size = 4, hjust = 0.5) +
  geom_text(data = ann_text_2,
            aes(x = week_date, y = deaths,
                label = "All hospital related deaths\nare much lower than\nthe historical average\nin recent weeks"),
            size = 4, hjust = 0.5) +
  geom_text(data = ann_text_3,
            aes(x = week_date, y = deaths,
                label = "Non Covid-19 related\nCare Home deaths\nare now lower than normal"),
            size = 4, hjust = 0.5) +
  geom_segment(data = ann_text_4,
               aes(x = ymd(20200608), xend = ymd(20200608), y = 300, yend = 380),
               size = 0.4,
               arrow = arrow(length = unit(0.075, "inches"), ends = "both")) +
  geom_text(data = ann_text_4, 
            aes(x = week_date, y = deaths,
                label = "The gap between historical\nhigh and both types of death\nat home remains high"),
            size = 4, hjust = 0.5) +
  scale_colour_wsj(labels = c("All deaths\n2020", 
                              "Non Covid-19 deaths\n2020"),
                      guide = guide_legend(label.hjust = 0.5,
                                           override.aes = list(
                                             size = 5))) +
  scale_y_continuous(limits = c(0, 900)) +
  scale_x_date(date_labels = "%b",
               breaks = scales::pretty_breaks(),
               limits = c(ymd(20191230), ymd(20201214))) +
  theme(legend.position = "top") +
  labs(x = "Week number",
       y = "Deaths(n)",
       colour = "",
       caption = glue::glue("Source:NRS Scotland\nExcluding other institutional place of death\ntotal n = {other_inst_all_2020} in 2020")) 
fig_10
```

```{r, eval=FALSE}
ggsave("plots/20200617_fig_10.png", fig_10, width = 14, height =10.5, dpi = 300)
```


## Fig extra 1

```{r}
fig_extra <- main_tab %>% 
  filter(location != "Other institution") %>% 
  mutate(diff = deaths_nonCovid_2020 - mean_deaths) %>% 
  ggplot(aes(week_date, diff, fill = location)) +
  geom_col() +
  facet_wrap(~location) +
  scale_fill_wsj() +
  scale_y_continuous(limits = c(-200, 300)) +
  theme(legend.position = "none") +
  labs(title = "Non-COVID-19 deaths by location in 2020",
       subtitle = "Difference from weekly average 2015-19", 
       x = "",
       y = "",
       caption = glue::glue("Source:NRS Scotland\nExcluding other institutional place of death total n = {other_inst_all_2020} in 2020\n\U00A9 David AG Henderson"))
fig_extra
```

```{r, eval=FALSE}
ggsave("plots/20200629_fig_extra.pdf", fig_extra, width = 12, height =9, dpi = 300)
```


## Fig extra 2

```{r, warning=FALSE, message=FALSE}
main_tab %>% 
  filter(location != "Other institution") %>% 
  mutate(diff = deaths_all_2020 - mean_deaths) %>% 
  ggplot(aes(week_date, diff, fill = location)) +
  geom_col() +
  facet_wrap(~location) +
  scale_fill_wsj() +
  scale_y_continuous(limits = c(-100, 500),
                     breaks = scales::pretty_breaks()) +
  theme(legend.position = "none") +
  labs(title = "All-cause deaths by location in 2020",
       subtitle = "Difference from weekly average 2015-19", 
       x = "",
       y = "",
       caption = glue::glue("Source:NRS Scotland\nExcluding other institutional place of death total n = {other_inst_all_2020} in 2020\n\U00A9 David AG Henderson")) -> fig_extra_2
fig_extra_2
```

```{r, eval=FALSE}
ggsave("plots/20200629_fig_extra_2.pdf", fig_extra_2, width = 12, height =9, dpi = 300)
```



## Fix extra 3

```{r}
fig_extra_3 <- main_tab %>% 
  filter(location != "Other institution") %>% 
  mutate(diff = deaths_nonCovid_2020 - mean_deaths,
         pct_change = round(diff/mean_deaths*100,1)) %>% 
  ggplot(aes(week_number, pct_change, fill = location)) +
  geom_col() +
  facet_wrap(~location) +
  scale_fill_wsj() +
  scale_y_continuous(breaks = scales::pretty_breaks(),
                     labels = scales::percent_format(scale = 1)) +
  theme(legend.position = "none") +
  labs(title = "Non-COVID-19 deaths by location in 2020",
       subtitle = "Percentage change from weekly average 2015-19", 
       x = "Week number",
       y = "",
       caption = glue::glue("Source:NRS Scotland\nExcluding other institutional place of death total n = {other_inst_all_2020} in 2020\n\U00A9 David AG Henderson"))
fig_extra_3
```

```{r, eval=FALSE}
ggsave("plots/20200629_fig_extra_3.pdf", fig_extra_3, width = 12, height =9, dpi = 300)
```

```{r}
fig_extra_4 <- main_tab %>% 
  filter(location != "Other institution") %>% 
  mutate(diff = deaths_all_2020 - mean_deaths,
         pct_change = round(diff/mean_deaths*100,1)) %>% 
  ggplot(aes(week_number, pct_change, fill = location)) +
  geom_col() +
  facet_wrap(~location) +
  scale_fill_wsj() +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  scale_y_continuous(limits = c(-50, 200),
                     breaks = scales::pretty_breaks(),
                     labels = scales::percent_format(scale = 1)) +
  theme(legend.position = "none") +
  labs(title = "All-cause deaths by location in 2020",
       subtitle = "Percentage change from weekly average 2015-19", 
       x = "Week number",
       y = "",
       caption = glue::glue("Source:NRS Scotland\nExcluding other institutional place of death total n = {other_inst_all_2020} in 2020\n\U00A9 David AG Henderson"))
fig_extra_4
```

```{r, eval=FALSE}
ggsave("plots/20200710_fig_extra_4.png", fig_extra_4, width = 12, height =9, dpi = 300)
```


```{r}
fig_extra_5 <-
  main_tab %>% 
  filter(location != "Other institution") %>% 
  ggplot(aes(week_number, deaths_covid_2020, fill = location)) +
  geom_col() +
  facet_wrap(~location) +
  scale_fill_wsj() +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  scale_y_continuous(limits = c(0, 400),
                     breaks = scales::pretty_breaks()) +
  theme(legend.position = "none") +
  labs(title = "COVID-19 deaths by location in 2020",
       subtitle = "", 
       x = "Week number",
       y = "",
       caption = glue::glue("Source:NRS Scotland\nExcluding other institutional place of death total n = {other_inst_all_2020} in 2020\n\U00A9 David AG Henderson"))
fig_extra_5
```


```{r, eval=FALSE}
ggsave("plots/20200710_fig_extra_5.png", fig_extra_5, width = 12, height =9, dpi = 300)
```

# Session Info

```{r}
devtools::session_info()
```


